Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-12-16 Manuscript/Manuscript.Rmd
Some suggestions to the intro

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Munoz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  Multilevel data is not spared the ubiquitous problem of missing information. This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
output: rticles::jss_article
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

Research into any field with a hierarchical or clustered nature of observations may yield multilevel data. In the typical case, individuals are nested within groups, but there are many different types of multilevel data. In the medical field, clustering occurs at e.g., the hospital or center level in registry data, or at the study-level in meta-analyses (IPDMA^[remove or write in full?]). In the social sciences and official statistics we can find clustering e.g. at the country-level, or as imposed by a multi-stage sampling design. For the sake of legibility, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit' throughout this paper.^[Alternatives to unit: participant/response/record.] And, for reasons of brevity, we only discuss clustering between units, not within units (e.g., timeseries or longitudinal data).^[Also, add that we'll only discuss two levels, not more?]

Analyzing multilevel data requires special care, compared to 'regular', single level data. For instance, there may be cross-level interactions between unit-level variables and cluster-level variables. The cluster to which a unit belongs may influence the unit-level observations, and vice versa for the the units that make up the cluster [@hox17]. These relations can and should be taken into account when developing analysis models for multilevel data.^[Explain ICC here? The percentage of variance attributed to the cluster-level is expressed by the intra-class coefficient (ICC). The ICC can also be interpreted as the expected correlation between two randomly sampled units in same cluster. So if the ICC is high, a lot of variability in a variable is due to the clustering, which should be modeled accordingly.] Multilevel models typically include separate intercepts for each cluster, which relieves one restriction imposed by single-level models: equal group means across clusters. Additionally, there may be random predictor effects and/or random error terms (residual error variances), see e.g. @hox17 and @jong21.^[Add that heterogeneity refers to variability within clusters.] There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

@@

Multilevel data is not spared the ubiquitous problem of missing information. Just as in single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for missingness at multiple levels. Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13. In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 3 clusters, and 3 variables of interest. Variable `X1` is completely observed. Variable `X2` is systematically missing, `X3` is sporadically missing.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters). In Figure 1, the former implies that the unobserved values for units 3 and 4 on variable `X2` would be equal. With the latter, the values may differ. Depending on the missing data pattern^[add missing data mechanisms here?], there are more or less optimal way of accommodating the missingness.

$$
\begin{matrix} 
                          &X1     & X2          & X3            \\
   \hline
   \mathrm{cluster 1}\{   &x_{11} &x_{12}       &\mathtt{NA}    \\
                          &x_{21} &x_{22}       &x_{32}         \\
   \mathrm{cluster 2}\{   &x_{31} &\mathtt{NA}  &x_{33}         \\
                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
   \mathrm{cluster 3}\{   &x_{51} &x_{52}       &x_{53}         \\
\end{matrix} 
$$

Ignoring missing data in research endeavors is almost never a good idea. Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power. Instead, the missingness should be accommodated *before* or *within* the analysis of scientific interest. Especially the former is very generic and popular. Imputing (i.e., filling in) the missing values splits the missing data problem from the scientific problem. The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which solves the missingness one variable at a time, iteratively. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.^[Discuss missingness mechanisms before this point, add references @yuce08 and @hox15.]

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Munoz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  Multilevel data is not spared the ubiquitous problem of missing information. This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: rticles::jss_article
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

~~Research into any field with a hierarchical or clustered nature of observations may yield multilevel data. In the typical case, individuals are nested within groups, but there are many different types of multilevel data. In the medical field, clustering occurs at e.g., the hospital or center level in registry data, or at the study-level in meta-analyses (IPDMA). In the social sciences and official statistics we can find clustering e.g. at the country-level, or as imposed by a multi-stage sampling design. For the sake of legibility, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit' throughout this paper. And, for reasons of brevity, we only discuss clustering between units, not within units (e.g., timeseries or longitudinal data). Also, add that we'll only discuss two levels, not more?~~

In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded. Ignoring such structures may be harmful to the inferences and can yield biased results. On the other hand, analyzing such multilevel data requires specialized techniques that take the clustered structure into account. Imagine a case where cross-level interactions between unit-level variables and cluster-level variables are present. The cluster to which a unit belongs may then influence the unit-level observations - and vice versa - for each of the units that make up the cluster [@hox17]. These relations can and should be taken into account when developing analysis models for multilevel data for the simple reason that groups of observations share some common variance.

The variability due to clustering is often measured by means of the intra-class coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure. Multilevel models typically accommodate for this variability by including separate intercepts for each cluster. Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters. Additionally, there may be random predictor effects and/or random error terms (residual error variances), see e.g. @hox17 and @jong21.^[Add that heterogeneity refers to variability within clusters.] There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

@@

The process of analyzing multilevel data is further complicated when not all observations are observed. Just as with single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for clustered missingness. Missingness in multilevel data can therefore be categorized into two general patterns: systematic missingness and sporadic missingness, see @resc13. We have visualized the difference between these types of missingness in Figure 1. The figure shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n=5$ units and $p=3$ columns distributed over 3 clusters. Column `X1` is completely observed. Column `X2` is systematically missing and column `X3` is sporadically missing.^[Explain why. GV: niet nodig]

Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters). In Figure 1, the former would imply that the unobserved values for units 3 and 4 on column `X2` are identical. With the latter, the values would differ. The optimal strategy for dealing with the missingness may therefore depend on the observed missing data pattern. ^[add missing data mechanisms here? GV: No, patterns are different from mechanisms]

$$
\begin{matrix} 
                          &X1     & X2          & X3            \\
   \hline
   \mathrm{cluster 1}\{ \ &x_{11} &x_{12}       &\mathtt{NA}    \\
                             \ &x_{21}  &x_{22}       &x_{32}         \\
   \mathrm{cluster 2}\{  &x_{31} &\mathtt{NA}  &x_{33}         \\
                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
   \mathrm{cluster 3}\{  &x_{51} &x_{52}       &x_{53}         \\
\end{matrix} 
$$
[GV Ik vind de latex tabel niet tof. ]

Ignoring the missingness in analyses can be extremely harmful to inferences. Complete case analysis (i.e., excluding all units with one or more missing entries) can introduce bias in statistical inference and lowers statistical power. Instead, the missingness should be accommodated *before* or *within* the analysis of scientific interest. Especially the former is very generic and popular ans is widely known as imputation. Imputing (i.e., filling in) the missing values seperates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analysed as if it were completely observed. The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. In this paper, we'll discuss how to use \pkg{mice} in the context of multilevel data, under varying missing data mechanisms.^[Discuss missingness mechanisms before this point, add references @yuce08 and @hox15.]

### 2021-12-16 Manuscript/Manuscript.tex
Some suggestions to the intro

\usepackage{graphicx}
\usepackage{mathtools}

@@

Research into any field with a hierarchical or clustered nature of
observations may yield multilevel data. In the typical case, individuals
are nested within groups, but there are many different types of
multilevel data. In the medical field, clustering occurs at e.g., the
hospital or center level in registry data, or at the study-level in
meta-analyses (IPDMA\footnote{remove or write in full?}). In the social
sciences and official statistics we can find clustering e.g.~at the
country-level, or as imposed by a multi-stage sampling design. For the
sake of legibility, we will refer to the grouping variable as `cluster',
and the grouped variable as `(sample) unit' throughout this
paper.\footnote{Alternatives to unit: participant/response/record.} And,
for reasons of brevity, we only discuss clustering between units, not
within units (e.g., timeseries or longitudinal data).\footnote{Also, add
  that we'll only discuss two levels, not more?}

Analyzing multilevel data requires special care, compared to `regular',
single level data. For instance, there may be cross-level interactions
between unit-level variables and cluster-level variables. The cluster to
which a unit belongs may influence the unit-level observations, and vice
versa for the the units that make up the cluster \citep{hox17}. These
relations can and should be taken into account when developing analysis
models for multilevel data.\footnote{Explain ICC here? The percentage of
  variance attributed to the cluster-level is expressed by the
  intra-class coefficient (ICC). The ICC can also be interpreted as the
  expected correlation between two randomly sampled units in same
  cluster. So if the ICC is high, a lot of variability in a variable is
  due to the clustering, which should be modeled accordingly.}
Multilevel models typically include separate intercepts for each
cluster, which relieves one restriction imposed by single-level models:
equal group means across clusters. Additionally, there may be random
predictor effects and/or random error terms (residual error variances),
see e.g. \citet{hox17} and \citet{jong21}.\footnote{Add that
  heterogeneity refers to variability within clusters.} There are many
names for models that take clustering into account. Some popular
examples are `multilevel models', `hierarchical models', `mixed effect
models' and `random effect models'.

@@

Multilevel data is not spared the ubiquitous problem of missing
information. Just as in single level data, missingness may occur at the
unit level. But with multiple levels of data comes the potential for
missingness at multiple levels. Missingness in multilevel data can be
categorized into two general patterns: systematical missingness and
sporadic missingness, see \citet{resc13}. In figure 1, we show a dataset
with units in the rows and variables in the columns, there are 5 units
nested within 3 clusters, and 3 variables of interest. Variable
\texttt{X1} is completely observed. Variable \texttt{X2} is
systematically missing, \texttt{X3} is sporadically missing.\footnote{Explain
  why.} Systematic missingness can be further subdivided into unobserved
constants (i.e., the same value within clusters) and non-measured random
variables (which may differ per unit within clusters). In Figure 1, the
former implies that the unobserved values for units 3 and 4 on variable
\texttt{X2} would be equal. With the latter, the values may differ.
Depending on the missing data pattern\footnote{add missing data
  mechanisms here?}, there are more or less optimal way of accommodating
the missingness.

\[
\begin{matrix} 
                          &X1     & X2          & X3            \\
   \hline
   \mathrm{cluster 1}\{   &x_{11} &x_{12}       &\mathtt{NA}    \\
                          &x_{21} &x_{22}       &x_{32}         \\
   \mathrm{cluster 2}\{   &x_{31} &\mathtt{NA}  &x_{33}         \\
                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
   \mathrm{cluster 3}\{   &x_{51} &x_{52}       &x_{53}         \\
\end{matrix} 
\]

Ignoring missing data in research endeavors is almost never a good idea.
Complete case analysis (i.e., excluding all units with one or more
missing entries) can introduce bias in statistical inference and lowers
statistical power. Instead, the missingness should be accommodated
\emph{before} or \emph{within} the analysis of scientific interest.
Especially the former is very generic and popular. Imputing (i.e.,
filling in) the missing values splits the missing data problem from the
scientific problem. The \proglang{R} package \pkg{mice} has become the
de-facto standard for imputation by chained equations, which solves the
missingness one variable at a time, iteratively. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
the context of multilevel data, under varying missing data
mechanisms.\footnote{Discuss missingness mechanisms before this point,
  add references \citet{yuce08} and \citet{hox15}.}

@@

Toy example from
\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{Heckman
Github repo}.

>>>>

\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{ulem}

@@

\sout{Research into any field with a hierarchical or clustered nature of
observations may yield multilevel data. In the typical case, individuals
are nested within groups, but there are many different types of
multilevel data. In the medical field, clustering occurs at e.g., the
hospital or center level in registry data, or at the study-level in
meta-analyses (IPDMA). In the social sciences and official statistics we
can find clustering e.g.~at the country-level, or as imposed by a
multi-stage sampling design. For the sake of legibility, we will refer
to the grouping variable as `cluster', and the grouped variable as
`(sample) unit' throughout this paper. And, for reasons of brevity, we
only discuss clustering between units, not within units (e.g.,
timeseries or longitudinal data). Also, add that we'll only discuss two
levels, not more?}

In many contemporary data analysis efforts, some form of hierarchical or
clustered structure is recorded. Ignoring such structures may be harmful
to the inferences and can yield biased results. On the other hand,
analyzing such multilevel data requires specialized techniques that take
the clustered structure into account. Imagine a case where cross-level
interactions between unit-level variables and cluster-level variables
are present. The cluster to which a unit belongs may then influence the
unit-level observations - and vice versa - for each of the units that
make up the cluster \citep{hox17}. These relations can and should be
taken into account when developing analysis models for multilevel data
for the simple reason that groups of observations share some common
variance.

The variability due to clustering is often measured by means of the
intra-class coefficient (ICC). The ICC can be seen as the percentage of
variance that can be attributed to the cluster-level, where a high ICC
would indicate that a lot of variability is due to the cluster
structure. Multilevel models typically accommodate for this variability
by including separate intercepts for each cluster. Such fixed effects
relieve the restriction imposed by single-level models: equal group
means across clusters. Additionally, there may be random predictor
effects and/or random error terms (residual error variances), see e.g.
\citet{hox17} and \citet{jong21}.\footnote{Add that heterogeneity refers
  to variability within clusters.} There are many names for models that
take clustering into account. Some popular examples are `multilevel
models', `hierarchical models', `mixed effect models' and `random effect
models'.

@@

The process of analyzing multilevel data is further complicated when not
all observations are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Missingness in
multilevel data can therefore be categorized into two general patterns:
systematic missingness and sporadic missingness, see \citet{resc13}. We
have visualized the difference between these types of missingness in
Figure 1. The figure shows an \(n \times p\) set
\(\mathbf{X} = X_1, \dots, X_p\), with \(n=5\) units and \(p=3\) columns
distributed over 3 clusters. Column \texttt{X1} is completely observed.
Column \texttt{X2} is systematically missing and column \texttt{X3} is
sporadically missing.\footnote{Explain why. GV: niet nodig}

Systematic missingness can be further subdivided into unobserved
constants (i.e., the same value within clusters) and non-measured random
variables (which may differ per unit within clusters). In Figure 1, the
former would imply that the unobserved values for units 3 and 4 on
column \texttt{X2} are identical. With the latter, the values would
differ. The optimal strategy for dealing with the missingness may
therefore depend on the observed missing data pattern. \footnote{add
  missing data mechanisms here? GV: No, patterns are different from
  mechanisms}

\[
\begin{matrix} 
                          &X1     & X2          & X3            \\
   \hline
   \mathrm{cluster 1}\{ \ &x_{11} &x_{12}       &\mathtt{NA}    \\
                             \ &x_{21}  &x_{22}       &x_{32}         \\
   \mathrm{cluster 2}\{  &x_{31} &\mathtt{NA}  &x_{33}         \\
                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
   \mathrm{cluster 3}\{  &x_{51} &x_{52}       &x_{53}         \\
\end{matrix} 
\] {[}GV Ik vind de latex tabel niet tof. {]}

Ignoring the missingness in analyses can be extremely harmful to
inferences. Complete case analysis (i.e., excluding all units with one
or more missing entries) can introduce bias in statistical inference and
lowers statistical power. Instead, the missingness should be
accommodated \emph{before} or \emph{within} the analysis of scientific
interest. Especially the former is very generic and popular ans is
widely known as imputation. Imputing (i.e., filling in) the missing
values seperates the missing data problem from the scientific problem:
missing data are replaced by plausible values whereafter the completed
data is analysed as if it were completely observed. The \proglang{R}
package \pkg{mice} has become the de-facto standard for imputation by
chained equations, which iteratively solves the missingness on a
variable-by-variable basis. \pkg{mice} is known to yield valid
inferences under many different missing data circumstances
\citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
the context of multilevel data, under varying missing data
mechanisms.\footnote{Discuss missingness mechanisms before this point,
  add references \citet{yuce08} and \citet{hox15}.}

@@

Toy example from
\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{Heckman
Github repo}. We will use the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{marital} Marital status,
\item
  \texttt{condom} Condom use during last intercourse,
\item
  \texttt{smoke} Smoker (levels).
\end{itemize}

\begin{center}\includegraphics{Manuscript_files/figure-latex/hiv-1} \end{center}

\end{CodeChunk}
